---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("06-01-2025","08-01-2025","10-01-2025") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", 
  month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction](lecture-01-08-2025.html).

`r i <- i + 1; d[i]` [Linear models, indicator variables](lecture-01-10-2025.html).

***

`r i <- i + 1; d[i]` [Introduction to R and the `lm` function for specifying linear models](lecture-01-13-2025.html).

`r i <- i + 1; d[i]` [Linear functions of model parameters, using the `lincon` function](lecture-01-15-2025.html).

`r i <- i + 1; d[i]` [Using the `contrast` function for inferences concerning linear functions of model parameters, plotting estimated expected responses with `ggplot`](lecture-01-17-2025.html). 
 
***

`r i <- i + 1; d[i]` **Holiday**. 

`r i <- i + 1; d[i]` [Inference for linear models](lecture-01-22-2025.html).

`r i <- i + 1; d[i]` [Inference for linear models --- continued](lecture-01-24-2025.html).
 
***

`r i <- i + 1; d[i]` [R workflow demonstration for linear models](lecture-01-27-2025.html).

`r i <- i + 1; d[i]` [Estimated expected response, prediction intervals, and some visualization](lecture-01-29-2025.html).

`r i <- i + 1; d[i]` [Marginal means](lecture-01-31-2025.html).
 
***

`r i <- i + 1; d[i]` [Modeling nonlinearity](lecture-02-03-2025.html).

`r i <- i + 1; d[i]` [Nonlinear regression](lecture-02-05-2025.html).

`r i <- i + 1; d[i]` [Nonlinear regression --- continued](lecture-02-07-2025.html).
 
***

`r i <- i + 1; d[i]` [Assumptions](lecture-02-10-2025.html).

`r i <- i + 1; d[i]` [Assumptions --- continued](lecture-02-12-2025.html).

`r i <- i + 1; d[i]` [Heteroscedasticity, weighted least squares, and iteratively weighted least squares](lecture-02-14-2025.html).
 
***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` [Iteratively weighted least squares --- continued, parametric models for heteroscedasticity, robust standard errors](lecture-02-19-2025.html).

`r i <- i + 1; d[i]` [In-class demonstration of nonlinear regression and dealing with heteroscedasticity](lecture-02-21-2025.html).
 
***

`r i <- i + 1; d[i]` [Introduction to Poisson regression and generalized linear models](lecture-02-24-2025.html). 

`r i <- i + 1; d[i]` [Poisson regression and rate ratios](lecture-02-26-2025.html).

`r i <- i + 1; d[i]` [Modeling rates with Poisson regression](lecture-02-28-2025.html).
 
***

`r i <- i + 1; d[i]` [Introduction to logistic regression and odds ratios](lecture-03-03-2025.html).

`r i <- i + 1; d[i]` [Logistic regression --- continued](lecture-03-05-2025.html).

`r i <- i + 1; d[i]` [Using the **emmeans** package, the relationship between Poisson and logistic regression, separation](lecture-03-07-2025.html).
 
***

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.
 
***

`r i <- i + 1; d[i]` [In-class demonstration of Poisson and logistic regression](lecture-03-17-2025.html).

`r i <- i + 1; d[i]` [Maximum likelihood](lecture-03-19-2025.html).

`r i <- i + 1; d[i]` [Over-dispersion](lecture-03-21-2025.html).
 
***

`r i <- i + 1; d[i]` [Negative binomial regression, robust standard errors, gamma regression](lecture-03-24-2025.html).

`r i <- i + 1; d[i]` [Alternative link functions](lecture-03-26-2025.html).

`r i <- i + 1; d[i]` [Marginal effects](lecture-03-28-2025.html).
 
***

`r i <- i + 1; d[i]` [The delta method](lecture-03-31-2025.html).

`r i <- i + 1; d[i]` [Introduction to survival analysis, accelerated failure time models](lecture-04-02-2025.html).
 
`r i <- i + 1; d[i]` [Censoring, survival functions](lecture-04-04-2025.html).
 
***

`r i <- i + 1; d[i]` [Hazard functions, proportional hazards models, plotting hazard and probability density functions](lecture-04-07-2025.html).

`r i <- i + 1; d[i]` [Proportional hazards models --- continued, semi-parametric proportional hazards models, discrete survival models](lecture-04-09-2025.html).

`r i <- i + 1; d[i]` [Introduction to categorical regression models, sequential models](lecture-04-11-2025.html).
 
***

`r i <- i + 1; d[i]` [Sequential and proportional odds models](lecture-04-14-2025.html).

`r i <- i + 1; d[i]` [Multinomial logit and discrete choice models](lecture-04-16-2025.html).

`r i <- i + 1; d[i]` [Introduction to the incidental parameter problem, generalized estimating equations](lecture-04-18-2025.html).
 
***

`r i <- i + 1; d[i]` [Fixed-effects approach to the incidental parameter problem](lecture-04-21-2025.html).

`r i <- i + 1; d[i]` [Random-effects approach to the incidental parameter problem](lecture-04-23-2025.html).

`r i <- i + 1; d[i]` [Nonlinear regression models with random effects, crossed and nested random effects](lecture-04-25-2025.html).
 
***

`r i <- i + 1; d[i]` [Prediction error, cross-validation, AIC](lecture-04-28-2025.html).

`r i <- i + 1; d[i]` [Generalized additive models](lecture-04-30-2025.html).

`r i <- i + 1; d[i]` 






