---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("06-01-2025","08-01-2025","10-01-2025") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", 
  month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction](lecture-01-08-2025.html).

`r i <- i + 1; d[i]` [Linear models, indicator variables](lecture-01-10-2025.html).

***

`r i <- i + 1; d[i]` [Introduction to R and the `lm` function for specifying linear models](lecture-01-13-2025.html).

`r i <- i + 1; d[i]` [Linear functions of model parameters, using the `lincon` function](lecture-01-15-2025.html).

`r i <- i + 1; d[i]` [Using the `contrast` function for inferences concerning linear functions of model parameters, plotting estimated expected responses with `ggplot`](lecture-01-17-2025.html). 
 
***

`r i <- i + 1; d[i]` **Holiday**. 

`r i <- i + 1; d[i]` [Inference for linear models](lecture-01-22-2025.html).

`r i <- i + 1; d[i]` [Inference for linear models --- continued](lecture-01-24-2025.html).
 
***

`r i <- i + 1; d[i]` [R workflow demonstration for linear models](lecture-01-27-2025.html).

`r i <- i + 1; d[i]` [Estimated expected response, prediction intervals, and some visualization](lecture-01-29-2025.html).

`r i <- i + 1; d[i]` [Marginal means](lecture-01-31-2025.html).
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]`  

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 






