---
output:
  word_document: default
  pdf_document: default
  html_document:
    theme: readable
---

```{r, echo = FALSE, message = FALSE}
library(lubridate)
x <- strsplit(this.path::this.path(), split = "/")[[1]]
x <- strsplit(x[length(x)], split = "-")[[1]]
date <- paste(x[2], x[3], year(Sys.Date()), sep = "-")
weekday <- wday(mdy(date), label = TRUE, abbr = FALSE)
month <- month(mdy(date), label = TRUE, abbr = FALSE)
day <- day(mdy(date))
```

---
title: `r paste(weekday, ", ", month, " ", day, sep = "")`
output:
  html_document: 
    theme: readable
  pdf_document: default
urlcolor: blue
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{array}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", message = FALSE, out.width = "100%", fig.align = "center", cache = FALSE, dev = ifelse(knitr::is_html_output(), "png", "pdf"))
```

```{r packages, echo = FALSE}
library(tidyverse)
suppressWarnings(library(kableExtra))
```

```{r utilities, echo = FALSE}
source("utilities.R")
```

```{r options, echo = FALSE}
options(digits = 3, width = 100)
```

`r ifelse(knitr::is_html_output(), paste("You can also download a [PDF](lecture-", date, ".pdf) copy of this lecture.", sep = ""), "")`


## Poisson Regression for Rates

The $i$-th observed *rate* $R_i$ can be written as
$$
	R_i = C_i/S_i,
$$
where $C_i$ is a *count* and $S_i$ is the "size" of the interval in which the counts are observed. Examples include fish per minute, epileptic episodes per day, or defects per (square) meter. In some cases $S_i$ is referred to as the "exposure" of the $i$-th observation. 

Assume that the count $C_i$ has a Poisson distribution and that
$$
  E(C_i) = S_i \underbrace{\exp(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_k x_{ik})}_{\lambda_i},
$$
where $\lambda_i$ is the expected count *per unit* (e.g., per minute) so that $S_i\lambda_i$ is then the expected count per $S_i$ (e.g., per hour if $S_i$ = 60, per day if $S_i$ = 1440, or per second if $S_i$ = 1/60). The expected *rate* is then
$$
  E(R_i) = E(C_i/S_i) = E(C_i)/S_i = \exp(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_k x_{ik}),
$$
if we treat exposure as *fixed* (like we do $x_{i1}, x_{i2}, \dots, x_{ik}$). But rather than using $R_i$ as the response variable we can use $C_i$ as the response variable in a Poisson regression model where
$$
  E(C_i) = S_i \exp(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_k x_{ik})
         =  \exp(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_k x_{ik} + \log S_i),
$$
and where $\log S_i$ is an "offset" variable (i.e., basically an explanatory variable where it's $\beta_j$ is "fixed" at one). 

Note: If $S_i$ is a *constant* for all observations so that $S_i = S$ then we can write the model as 
$$
  E(C_i) = \exp(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_k x_{ik} + \log S_i) =
  \exp(\beta_0^* + \beta_1 x_{i1} + \beta_2 x_{i2} + \cdots + \beta_k x_{ik}),
$$
where $\beta_0^* = \log(S) + \beta_0$ so that the offset is "absorbed" into $\beta_0$, and we do not need to be concerned about it. Including an offset is only necessary if $S_i$ is not the same for all observations.

## Variance of Rates

Using rates as response variables in a linear or nonlinear model without accounting for $S_i$ is not advisable because of heteroscedasticity due to unequal $S_i$.

We have that $E(R_i) = E(C_i)/S_i$. But 
$$
  \mbox{Var}(R_i) = \mbox{Var}(C_i/S_i) = \mbox{Var}(C_i)/S_i^2 = E(R_i)S_i/S_i^2 = E(R_i)/S_i,
$$
because (a) $\mbox{Var}(Y/c) = \mbox{Var}(Y)/c^2$ if $c$ is a constant, $\mbox{Var}(C_i) = E(C_i)$ because $C_i$ has a Poisson distribution, and thus $E(C_i) = E(R_i)S_i$. Thus the variance of the rate depends on the expected response and $S_i$ (so larger/smaller $S_i$, then smaller/larger variance of $R_i)$. 

We can deal with this heteroscedasticity by either (a) using an appropriate offset variable in Poisson regression or a related model or (b) using weights of $w_i = S_i/E(R_i)$ in an iteratively weighted least squares with weights of $w_i = S_i/\hat{y}_i$. 

## Modeling Rates with Poisson Regression

Software for GLMs (and sometimes linear models) will often permit specification of an offset variable. In R this is done using `offset` in the model formula. 

**Example**: Consider the following data from an observational study of auto accidents. 
```{r, message = FALSE}
library(trtools)
head(accidents)
p <- ggplot(accidents, aes(x = location, y = accidents/years)) +
  geom_point(aes(size = years, color = treatment)) + 
  labs(x = "Location", y = "Accidents per Year", 
    size = "Years", color = "Treatment") + theme_minimal()
plot(p)
m <- glm(accidents ~ location + treatment + offset(log(years)), 
  data = accidents, family = poisson)
cbind(summary(m)$coefficients, confint(m))
exp(cbind(coef(m), confint(m)))
```
When using other tools like `contrast` or functions from the **emmeans** package, be sure to specify the offset. Typically we would use a value of one corresponding to one unit of whatever the offset represents (e.g., space or time). Here are the rate ratios for the treatment.
```{r}
trtools::contrast(m,
  a = list(treatment = "before", location = letters[1:8], years = 1),
  b = list(treatment = "after", location = letters[1:8], years = 1),
  cnames = letters[1:8], tf = exp)
trtools::contrast(m,
  a = list(treatment = "after", location = letters[1:8], years = 1),
  b = list(treatment = "before", location = letters[1:8], years = 1),
  cnames = letters[1:8], tf = exp)
```
Here are the estimated expected number of accidents per year at location `a`.
```{r}
trtools::contrast(m, a = list(treatment = c("before","after"), location = "a", years = 1),
  cnames = c("before","after"), tf = exp)
```
Here are the estimated expected number of accidents *per decade* at location `a`.
```{r}
trtools::contrast(m, a = list(treatment = c("before","after"), location = "a", years = 10),
  cnames = c("before","after"), tf = exp)
```
When using functions from the **emmeans** package we use the `offset` argument with the value specified on the log scale. Here are the estimated number of accidents per decade.
```{r}
library(emmeans)
emmeans(m, ~treatment|location, type = "response", offset = log(10))
```
Here is the rate ratio for the effect of treatment.
```{r}
pairs(emmeans(m, ~treatment|location, type = "response", offset = log(10)), infer = TRUE)
```
Use `reverse = TRUE` to "flip" the rate ratio. Also for rate ratios the size of the offset does not matter since it "cancels-out" in the ratio. Also since there is no interaction in this model which means the rate ratio does not depend on location, we can omit it when using `emmeans` (but not `contrast`).
```{r}
pairs(emmeans(m, ~treatment, type = "response"), infer = TRUE)
```
When using `predict` we need to be sure to also include the offset amount. Again, we would use a value of one assuming we want the number of events per unit space/time.
```{r}
d <- expand.grid(treatment = c("before","after"), location = letters[1:8], years = 1)
d$yhat <- predict(m, newdata = d, type = "response")
head(d)
p <- ggplot(accidents, aes(x = location, y = accidents/years)) +
  geom_point(aes(size = years, color = treatment)) + 
  labs(x = "Location", y = "Accidents per Year", 
    size = "Years", color = "Treatment") + theme_minimal() + 
  geom_point(aes(y = yhat, color = treatment), data = d) + 
  geom_line(aes(y = yhat, group = treatment, color = treatment), data = d)
plot(p)
```
We can use the `glmint` function from the **trtools** package if we want to produce confidence intervals for plots.
```{r}
d <- expand.grid(treatment = c("before","after"), location = letters[1:8], years = 1)
d$yhat <- predict(m, newdata = d, type = "response")
glmint(m, newdata = d)
cbind(d, glmint(m, newdata = d))
d <- cbind(d, glmint(m, newdata = d))
p <- ggplot(accidents, aes(x = location)) + 
  geom_point(aes(y = accidents/years, size = years), shape = 21, fill = "white") + 
  facet_wrap(~ treatment) + theme_minimal() + 
  labs(x = "Location", y = "Accidents per Year", size = "Years") + 
  geom_errorbar(aes(ymin = low, ymax = upp), data = d, width = 0.5) + 
  geom_point(aes(y = fit), data = d)
plot(p)
p <- ggplot(accidents, aes(x = location, color = treatment)) + 
  geom_point(aes(y = accidents/years, size = years), 
    position = position_dodge(width = 0.6), shape = 21, fill = "white") + 
  labs(x = "Location", y = "Accidents per Year", 
    size = "Years", color = "Treatment") + theme_minimal() + 
  geom_errorbar(aes(ymin = low, ymax = upp), data = d, 
    position = position_dodge(width = 0.6), width = 0.5) + 
  geom_point(aes(y = fit), data = d, position = position_dodge(width = 0.6))
plot(p)
```

**Example**: Consider the following data from an observational study that investigated the possible effect of the development of a commercial fishery on deep sea fish abundance. The figure below shows the number of fish per square meter of swept area from 147 trawls by mean depth in meters, and by whether the trawl was during one of two periods. The 1977-1989 period was from before the development of a commercial fishery, and the period 2000-2002 was when the fishery was active.
```{r, warning = FALSE}
library(COUNT)
data(fishing)
head(fishing)
p <- ggplot(fishing, aes(x = meandepth, y = totabund/sweptarea)) +
  geom_point(alpha = 0.5) + facet_wrap(~ period) + theme_minimal() +
  labs(x = "Mean Trawl Depth (meters)",
    y = "Fish Caught Per Square Meter Trawled")
plot(p)
```
An appropriate model for these data might be as follows.
```{r, warning = FALSE}
m <- glm(totabund ~ period * meandepth + offset(log(sweptarea)),
  family = poisson, data = fishing)
summary(m)$coefficients
d <- expand.grid(sweptarea = 1, period = c("1977-1989","2000-2002"), 
  meandepth = seq(800, 5000, length = 100))
d$yhat <- predict(m, newdata = d, type = "response")

p <- p + geom_line(aes(y = yhat), data = d)
plot(p)
```
What is the expected number of fish per square meter in 1977-1989 at depths of 1000, 2000, 3000, 4000, and 5000 meters? What is it in 2000-2002?
```{r}
trtools::contrast(m, 
  a = list(sweptarea = 1, 
    meandepth = c(1000,2000,3000,4000,5000), period = "1977-1989"),
  cnames = c("1000m","2000m","3000m","4000m","5000m"), tf = exp)
trtools::contrast(m, 
  a = list(sweptarea = 1,
    meandepth = c(1000,2000,3000,4000,5000), period = "2000-2002"),
  cnames = c("1000m","2000m","3000m","4000m","5000m"), tf = exp)
```
Here is how we can do that with **emmeans**.
```{r}
library(emmeans)
emmeans(m, ~meandepth|period, at = list(meandepth = seq(1000, 5000, by = 1000)),
  type = "response", offset = log(1))
```
Note that we can change the units of swept area very easily here. There are 10,000 square meters in a hectare. Here are the expected number of fish per hectare. 
```{r}
trtools::contrast(m, 
  a = list(sweptarea = 10000,
    meandepth = c(1000,2000,3000,4000,5000), period = "1977-1989"),
  cnames = c("1000m","2000m","3000m","4000m","5000m"), tf = exp)
trtools::contrast(m, 
  a = list(sweptarea = 10000,
    meandepth = c(1000,2000,3000,4000,5000), period = "2000-2002"),
  cnames = c("1000m","2000m","3000m","4000m","5000m"), tf = exp)
emmeans(m, ~meandepth|period, at = list(meandepth = seq(1000, 5000, by = 1000)),
  type = "response", offset = log(10000))
```
What is the rate ratio of fish per square meter in 2000-2002 versus 1977-1989 at 1000, 2000, 3000, 4000, and 5000 meters?
```{r}
trtools::contrast(m, 
  a = list(sweptarea = 1, meandepth = c(1000,2000,3000,4000,5000), period = "2000-2002"),
  b = list(sweptarea = 1, meandepth = c(1000,2000,3000,4000,5000), period = "1977-1989"),
  cnames = c("1000m","2000m","3000m","4000m","5000m"), tf = exp)
```
Here it is for 1977-1989 versus 2000-2002.
```{r}
trtools::contrast(m, 
  a = list(sweptarea = 1, meandepth = c(1000,2000,3000,4000,5000), period = "1977-1989"),
  b = list(sweptarea = 1, meandepth = c(1000,2000,3000,4000,5000), period = "2000-2002"),
  cnames = c("1000m","2000m","3000m","4000m","5000m"), tf = exp)
```
Now using **emmeans**.
```{r}
pairs(emmeans(m, ~meandepth*period, at = list(meandepth = seq(1000, 5000, by = 1000)),
  type = "response", offset = log(1)), by = "meandepth", infer = TRUE)
```
How does the expected number of fish per square meter change per 1000m of depth?
```{r}
# increasing depth by 1000m
trtools::contrast(m,
  a = list(sweptarea = 1, meandepth = 2000, period = c("1977-1989","2000-2002")),
  b = list(sweptarea = 1, meandepth = 1000, period = c("1977-1989","2000-2002")),
  cnames = c("1977-1989","2000-2002"), tf = exp)
# decreasing depth by 1000m
trtools::contrast(m,
  a = list(sweptarea = 1, meandepth = 1000, period = c("1977-1989","2000-2002")),
  b = list(sweptarea = 1, meandepth = 2000, period = c("1977-1989","2000-2002")),
  cnames = c("1977-1989","2000-2002"), tf = exp)
```
Here is how to do the latter with **emmeans**.
```{r}
pairs(emmeans(m, ~meandepth*period, at = list(meandepth = c(1000,2000)),
  offset = log(1), type = "response"), by = "period", infer = TRUE)
```

## Standardized Mortality Ratios

In epidemiology, the *standardized mortality ratio* (SMR) is the ratio of the *observed* number of deaths and the (estimated) *expected* number of deaths. Poisson regression with an offset can be used to model the SMR to determine if the number of deaths tends to be higher or lower than we would expect.

**Example**: Here is an example of an observational study using a Poisson regression model to investigate the relationship between lung cancer and radon exposure in counties in Minnesota. 

Note: The data manipulation and plotting is quite a bit more complicated than what you will normally see in this class, but I have included it in case you might be interested to see the code.

First we will process the data containing the observed and expected number of deaths due to lung cancer, where the latter are based on the known distribution of age and gender in the county. 
```{r}
lung <- read.table("http://faculty.washington.edu/jonno/book/MNlung.txt", 
  header = TRUE, sep = "\t") %>% 
  mutate(obs = obs.M + obs.F, exp = exp.M + exp.F) %>% 
  dplyr::select(X, County, obs, exp) %>%
  rename(county = County) %>% 
  mutate(county = tolower(county)) %>% 
  mutate(county = ifelse(county == "red", "red lake", county))
head(lung)
```
Now we will read in data to estimate the average radon exposure of residents of each county.
```{r}
radon <- read.table("http://faculty.washington.edu/jonno/book/MNradon.txt",
  header = TRUE) %>% group_by(county) %>% 
  summarize(radon = mean(radon)) %>% rename(X = county)
head(radon)
```
Next we merge the two data frames.
```{r}
radon <- left_join(lung, radon) %>% dplyr::select(-X)
head(radon)
```
For fun we can make some plots of the data by county.
```{r}
library(maps)
dstate <- map_data("state") %>% 
  filter(region == "minnesota")
dcounty <- map_data("county") %>% 
  filter(region == "minnesota") %>% 
  rename(county = subregion)
dcounty <- left_join(dcounty, radon) %>% 
  mutate(smr = obs/exp)
```

```{r}
no_axes <- theme_minimal() + theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank()
)

p <- ggplot(dcounty, aes(x = long, y = lat, group = group)) + coord_fixed(1.3) +
  geom_polygon(aes(fill = exp), color = "black", linewidth = 0.25) + 
  scale_fill_gradient(low = grey(0.95), high = grey(0.25), 
    trans = "log10", na.value = "pink") + 
  theme(legend.position = "inside", legend.position.inside = c(0.8,0.4)) + 
  no_axes + ggtitle("Expected Number of Cases") + labs(fill = "Cases")

plot(p)

p <- ggplot(dcounty, aes(x = long, y = lat, group = group)) + coord_fixed(1.3) +
  geom_polygon(aes(fill = smr), color = "black", linewidth = 0.25) + 
  scale_fill_gradient(low = grey(0.95), high = grey(0.25), na.value = "pink") +
  theme(legend.position = "inside", legend.position.inside = c(0.8,0.4)) + 
  no_axes + ggtitle("Standardized Mortality Ratio") + labs(fill = "SMR")

plot(p)

p <- ggplot(dcounty, aes(x = long, y = lat, group = group)) + coord_fixed(1.3) + 
  geom_polygon(aes(fill = radon), color = "black", linewidth = 0.25) + 
  scale_fill_gradient(low = grey(0.95), high = grey(0.25), na.value = "pink") + 
  theme(legend.position = "inside", legend.position.inside = c(0.8,0.4)) + 
  no_axes + ggtitle("Average Radon (pCi/liter)") + labs(fill = "Radon")

plot(p)
```
How does the expected SMR relate to radon? Consider the Poisson regression model
$$
  \log E(Y_i/E_i) = \beta_0 + \beta_1r_i,
$$
where $Y_i$ and $E_i$ are the observed and expected number of lung cancer deaths (or cases), respectively, in the $i$-th county, and $r_i$ is the average radon exposure in the $i$-th county. Here $Y_i/E_i$ is the SMR for the $i$-th county. We can also write this model as
$$
  \log E(Y_i) = \log E_i  + \beta_0 + \beta_1r_i,
$$
so $\log E_i$ is an *offset*.
```{r}
m <- glm(obs ~ offset(log(exp)) + radon, family = poisson, data = dcounty)
summary(m)$coefficients
exp(cbind(coef(m), confint(m)))
```
We should be careful and remember the [ecological fallacy](https://en.wikipedia.org/wiki/Ecological_fallacy) which states that relationships at the group level (e.g., county) do not necessarily hold at the individual level. Radon may be related to other variables (e.g., smoking) that affect the risk of lung cancer.
