---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("08-01-2024","10-01-2024","23-01-2024") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", 
  month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction](lecture-01-10-2024.html).

`r i <- i + 1; d[i]` [Linear models](lecture-01-12-2024.html). 

***

`r i <- i + 1; d[i]` **Holiday**. 

`r i <- i + 1; d[i]` [Introduction to R and the `lm` function for specifying a linear model](lecture-01-17-2024.html). 

`r i <- i + 1; d[i]` [Specifying linear models --- continued, linear combinations of parameters](lecture-01-19-2024.html).
 
***

`r i <- i + 1; d[i]` [Using the `contrast` function, plotting data and models](lecture-01-22-2024.html). 

`r i <- i + 1; d[i]` [Statistical inference for linear models](lecture-01-24-2024.html). 

`r i <- i + 1; d[i]` [Statistical inference for linear models --- continued](lecture-01-26-2024.html). 
 
***

`r i <- i + 1; d[i]` [R workflow demonstration for linear models](lecture-01-29-2024.html).

`r i <- i + 1; d[i]` [Estimated expected response, prediction intervals, and some visualization](lecture-01-31-2024.html).

`r i <- i + 1; d[i]` [Estimating marginal means](lecture-02-02-2024.html). 
 
***

`r i <- i + 1; d[i]` [Modeling nonlinear relationships with linear models](lecture-02-05-2024.html).

`r i <- i + 1; d[i]` [Introduction to nonlinear regression](lecture-02-07-2024.html). 

`r i <- i + 1; d[i]` [Nonlinear regression --- continued](lecture-02-09-2024.html).
 
***

`r i <- i + 1; d[i]` [Assumptions, residuals, and heteroscedasticity](lecture-02-12-2024.html).

`r i <- i + 1; d[i]` [Assumptions --- continued](lecture-02-14-2024.html).

`r i <- i + 1; d[i]` [Solutions for heteroscedasticity --- variance stabilizing transformations, weighted least squares](lecture-02-16-2024.html).
 
***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` [Iteratively weighted least squares, parametric models for heteroscedasticity, heteroscedastic consistent standard errors](lecture-02-21-2024.html).

`r i <- i + 1; d[i]` [R workflow demonstration for nonlinear regression and heteroscedasticity](lecture-02-23-2024.html).
 
***

`r i <- i + 1; d[i]` [Introduction to Poisson regression and generalized linear models](lecture-02-26-2024.html).

`r i <- i + 1; d[i]` [Interpretation of GLMs with log link functions, rate ratios](lecture-02-28-2024.html).

`r i <- i + 1; d[i]` [Using an offset variable with Poisson regression](lecture-03-01-2024.html).
 
***

`r i <- i + 1; d[i]` [Introduction to logistic regression](lecture-03-04-2024.html). 

`r i <- i + 1; d[i]` [Logistic regression --- continued, odds ratios](lecture-03-06-2024.html).

`r i <- i + 1; d[i]` [Using the **emmeans** package with Poisson and logistic regression models, the relationship between Poisson and logistic regression, separation](lecture-03-08-2024.html).
 
***

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 






