---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("09-01-2023","11-01-2023","13-01-2023") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", 
  month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction](lecture-01-11-2023.html).

`r i <- i + 1; d[i]` [Linear models and their parameterization](lecture-01-13-2023.html).

***

`r i <- i + 1; d[i]` **Holiday**. 

`r i <- i + 1; d[i]` [Introduction to linear model specification in R using the `lm` function](lecture-01-18-2023.html).

`r i <- i + 1; d[i]` [Linear models with both quantitative and categorical explanatory variables, linear combinations of parameters](lecture-01-20-2023.html).
 
***

`r i <- i + 1; d[i]` [Using the `contrast` function, plotting with `ggplot`](lecture-01-23-2023.html).

`r i <- i + 1; d[i]` [Introduction/review of statistical inference](lecture-01-25-2023.html). 

`r i <- i + 1; d[i]` [Introduction/review of statistical inference --- continued](lecture-01-27-2023.html).
 
***

`r i <- i + 1; d[i]` [R demonstration on linear models and using `lincon` and `contrast` with linear models](lecture-01-30-2023.html). The [first homework assignment](hw1.html) is now available.

`r i <- i + 1; d[i]` [Estimated expected responses, prediction, and visualization](lecture-02-01-2023.html).

`r i <- i + 1; d[i]` [Marginal means, introduction to the **emmeans** package](lecture-02-03-2023.html). 
 
***

`r i <- i + 1; d[i]` [Modeling nonlinear relationships with linear models](lecture-02-06-2023.html).

`r i <- i + 1; d[i]` [Introduction to nonlinear regression](lecture-02-08-2023.html).

`r i <- i + 1; d[i]` [Nonlinear regression examples](lecture-02-10-2023.html). 
 
***

`r i <- i + 1; d[i]` [Assumptions of (non)linear regression models](lecture-02-13-2023.html).

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]`  

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]`






