---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("09-01-2023","11-01-2023","13-01-2023") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", 
  month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction](lecture-01-11-2023.html).

`r i <- i + 1; d[i]` [Linear models and their parameterization](lecture-01-13-2023.html).

***

`r i <- i + 1; d[i]` **Holiday**. 

`r i <- i + 1; d[i]` [Introduction to linear model specification in R using the `lm` function](lecture-01-18-2023.html).

`r i <- i + 1; d[i]` [Linear models with both quantitative and categorical explanatory variables, linear combinations of parameters](lecture-01-20-2023.html).
 
***

`r i <- i + 1; d[i]` [Using the `contrast` function, plotting with `ggplot`](lecture-01-23-2023.html).

`r i <- i + 1; d[i]` [Introduction/review of statistical inference](lecture-01-25-2023.html). 

`r i <- i + 1; d[i]` [Introduction/review of statistical inference --- continued](lecture-01-27-2023.html).
 
***

`r i <- i + 1; d[i]` [R demonstration on linear models and using `lincon` and `contrast` with linear models](lecture-01-30-2023.html). The [first homework assignment](hw1.html) is now available.

`r i <- i + 1; d[i]` [Estimated expected responses, prediction, and visualization](lecture-02-01-2023.html).

`r i <- i + 1; d[i]` [Marginal means, introduction to the **emmeans** package](lecture-02-03-2023.html). 
 
***

`r i <- i + 1; d[i]` [Modeling nonlinear relationships with linear models](lecture-02-06-2023.html).

`r i <- i + 1; d[i]` [Introduction to nonlinear regression](lecture-02-08-2023.html).

`r i <- i + 1; d[i]` [Nonlinear regression examples](lecture-02-10-2023.html). 
 
***

`r i <- i + 1; d[i]` [Assumptions of (non)linear regression models](lecture-02-13-2023.html).

`r i <- i + 1; d[i]` [Assumptions --- continued](lecture-02-15-2023.html). The [solutions](hw1-solutions.html) to the first homework assignment are now available.

`r i <- i + 1; d[i]` [Dealing with heteroscedasticity --- transformations and weighted least squares](lecture-02-17-2023.html).
 
***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` [Dealing with heteroscedasticity --- iteratively weighted least squares, parametric models, and robust standard errors](lecture-02-22-2023.html).

`r i <- i + 1; d[i]` [R demonstration on nonlinear regression and weighted least squares](lecture-02-24-2023.html).
 
***

`r i <- i + 1; d[i]` [Introduction to Poisson regression](lecture-02-27-2023.html). The [second homework assignment](hw2.html) is now available.

`r i <- i + 1; d[i]` [Interpretation of Poisson regression models using rate ratios](lecture-03-01-2023.html).

`r i <- i + 1; d[i]` [Modeling rates by using an offset variable](lecture-03-03-2023.html).
 
***

`r i <- i + 1; d[i]` [Introduction to logistic regression](lecture-03-06-2023.html). 

`r i <- i + 1; d[i]` [Examples of logistic regression and use of odds ratios](lecture-03-08-2023.html). Second homework assignment due.

`r i <- i + 1; d[i]` [Using the **emmeans** package with Poisson and logistic regression, the relationship between Poisson and logistic regression, and separation and infinite parameter estimates](lecture-03-10-2023.html).
 
***

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.
 
***

`r i <- i + 1; d[i]` [Maximum likelihood, likelihood-based tests and confidence intervals, Wald tests and confidence intervals](lecture-03-20-2023.html). The [solutions](hw2-solutions.html) for the second homework assignment are now available.

`r i <- i + 1; d[i]` [Over-dispersion and quasi-likelihood](lecture-03-22-2023.html). 

`r i <- i + 1; d[i]` [R demonstration of Poisson and logistic regression](lecture-03-24-2023.html). The [third homework assignment](hw3.html) is now available. 
 
***

`r i <- i + 1; d[i]` [Distributions for over-dispersion, gamma and inverse-gaussian generalized linear models](lecture-03-27-2023.html).

`r i <- i + 1; d[i]` [Alternative link functions](lecture-03-29-2023.html).

`r i <- i + 1; d[i]` [Marginal effects](lecture-03-31-2023.html). 
 
***

`r i <- i + 1; d[i]` [The delta method](lecture-04-03-2023.html).

`r i <- i + 1; d[i]` [Introduction to survival analysis, accelerated failure time models](lecture-04-05-2023.html).
 
`r i <- i + 1; d[i]` [Censoring specification, interval censoring, and survival functions](lecture-04-07-2023.html). 
 
***

`r i <- i + 1; d[i]` [Hazard functions, proportional hazards models](lecture-04-10-2023.html). The [solutions for the third homework assignment](hw3-solutions.html) are now available.

`r i <- i + 1; d[i]` [Proportional hazards models --- continued, discrete survival time models](lecture-04-12-2023.html).

`r i <- i + 1; d[i]` [Sequential categorical regression models](lecture-04-14-2023.html). The [fourth homework assignment](hw4.html) is now available.
 
***

`r i <- i + 1; d[i]` [Sequential models --- continued, proportional odds models](lecture-04-17-2023.html).

`r i <- i + 1; d[i]` [Multinomial logit models](lecture-04-19-2023.html). 

`r i <- i + 1; d[i]` [The incidental parameter problem, marginal models and generalized estimating equations](lecture-04-21-2023.html).
 
***

`r i <- i + 1; d[i]` [Fixed-effects models](lecture-04-24-2023.html). 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Sixth homework assignment assigned, due by noon on Monday, May 15th**. 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]`






