---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("10-01-2022","12-01-2022","14-01-2022") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", 
  month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction](lecture-01-12-2022.html). 

`r i <- i + 1; d[i]` [Linear (regression) models](lecture-01-14-2022.html).

***

`r i <- i + 1; d[i]` **Holiday**. 

`r i <- i + 1; d[i]` [Introduction to linear model specification in R using the `lm` function](lecture-01-19-2022.html).

`r i <- i + 1; d[i]` [Linear models with quantitative and categorical explanatory variables, linear combinations of parameters](lecture-01-21-2022.html).
 
***

`r i <- i + 1; d[i]` [Contrasts and estimated expected responses, and plotting estimated expected responses](lecture-01-24-2022.html).

`r i <- i + 1; d[i]` [Least squares estimators, sampling distributions, standard errors, confidence intervals, and tests](lecture-01-26-2022.html).

`r i <- i + 1; d[i]` [Confidence intervals and tests, testing joint hypotheses, analysis of variance calculations](lecture-01-28-2022.html).
 
***

`r i <- i + 1; d[i]` [In-class exercise on specifying and using linear models](lecture-01-31-2022.html). The first [homework assignment](hw1.html) is now available.

`r i <- i + 1; d[i]` [Estimated expected responses, prediction, and visualization](lecture-02-02-2022.html).

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]`
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 




