---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("12-01-2026","14-01-2026","16-01-2026") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = FALSE), ", ", 
  month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 1
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction](lecture-01-14-2026.html).

`r i <- i + 1; d[i]` [Linear models](lecture-01-16-2026.html).

***

`r i <- i + 1; d[i]` **Holiday**. 

`r i <- i + 1; d[i]` [Introduction to R and the `lm` function for specifying linear models](lecture-01-21-2026.html).

`r i <- i + 1; d[i]` [Linear functions of the parameters of a linear model](lecture-01-23-2026.html).
 
***

`r i <- i + 1; d[i]` [Using the `contrast` function for inferences concerning linear functions of model parameters, plotting estimated expected responses with `ggplot`](lecture-01-26-2026.html).

`r i <- i + 1; d[i]` [Inference for linear models](lecture-01-28-2026.html).

`r i <- i + 1; d[i]` [Inference for linear models --- continued](lecture-01-30-2026.html).
 
***

`r i <- i + 1; d[i]` [In-class demonstration of R workflow with specification and inference for linear models](lecture-02-02-2026.html).

`r i <- i + 1; d[i]` [Inferences w/visualization for expected responses and predictions](lecture-02-04-2026.html).

`r i <- i + 1; d[i]` [Marginal means](lecture-02-06-2026.html).
 
***

`r i <- i + 1; d[i]` [Modeling nonlinearity (with linear models)](lecture-02-09-2026.html).

`r i <- i + 1; d[i]` [Introduction to nonlinear regression](lecture-02-11-2026.html).

`r i <- i + 1; d[i]` [Nonlinear regression --- continued](lecture-02-13-2026.html).
 
***

`r i <- i + 1; d[i]` **Holiday**.

`r i <- i + 1; d[i]` [Assumptions](lecture-02-18-2026.html).

`r i <- i + 1; d[i]` [Assumptions --- continued](lecture-02-20-2026.html).

***

`r i <- i + 1; d[i]` [Solutions for heteroscedasticity: variance-stabilizing transformations, weighted least squares](lecture-02-23-2026.html).

`r i <- i + 1; d[i]` [Iteratively weighted least squares, parametric models for heteroscedasticity, robust standard errors](lecture-02-25-2026.html).

`r i <- i + 1; d[i]` [In-class demonstration of nonlinear regression and dealing with heteroscedasticity](lecture-02-27-2026.html).
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.

`r i <- i + 1; d[i]` **Spring recess**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 






